Weighted Model Counting with Conditional Weights

### 2020-08-07 paper/paper.tex
The abstract + grammar

\begin{abstract} % TODO
%1. Introduction. In one sentence, what’s the topic?
  % WMC
%2. State the problem you tackle.
  % Not using additional literals
%3. Summarize (in one sentence) why nobody else has adequately answered the research question yet.
  % The dominant definition of WMC severely restricts what measures can be represented
%4. Explain, in one sentence, how you tackled the research question.
  % Defining weights akin to conditional probabilities.
%5. In one sentence, how did you go about doing the research that follows from your big idea.
  % ADDMC + Bayesian networks
%6. As a single sentence, what’s the key impact of your research?
  % Big improvements in inference speed that can be applied to other algorithms
  % and application domains.
\end{abstract}

@@

\Cref{sec:prelims} describes three equivalent perspectives on the subject based
on logic, set theory, and Boolean algebras. Furthermore, we describe the space
of functions on Boolean algebras and various operations on those functions. In
\cref{sec:wmc_as_measure}, WMC is introduced as the problem of computing the
value of a measure on a Boolean algebra. We show that not all measures can be
represented using literal-based WMC, but all Boolean algebras can be extended to
make any measure representable in such a manner.

@@

Finally, in \cref{sec:experiments}, we show how the new encoding results in
significantly improved inference speed on most instances, often by one or two
orders of magnitude. We explain the performance benefits by showing how the new
encoding has asymptotically fewer variables and ADDs.

@@

\paragraph{ADDs and their use in probabilistic inference.} ADDs provide an
efficient way to manipulate functions from a Boolean algebra to any algebraic
structure (most commonly the real numbers)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. They have been used to represent value
functions in Markov decision processes \cite{DBLP:conf/uai/HoeySHB99} and, for
Bayesian network inference, to represent each conditional probability table
(CPT) as an ADD \cite{DBLP:conf/icml/ZhaoMP15} as well as by combining ADDs and
arithmetic circuits into a  single representation
\cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous in this
situation because of their ability to fully exploit context-specific
independence, i.e., observable structure within a CPT that is not inherited from
the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

@@

\subsection{The Space of Functions on Boolean Algebras}
% TODO: clarify why this is needed

@@

Note that in this case we chose to interpret $\nu_b$ as $\Pr(b \mid a)$
while---with a different definition of $\nu_b$ that represents the joint
probability distribution $\Pr(a, b)$---$\nu_b$ by itself could induce $\mu$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage \cite{DBLP:conf/aaai/DudekPV20}.

@@

Second, even though the function $\phi$ produced by \cref{alg:encoding} is
constructed to have $2^U$ as its domain, sometimes the domain is effectively
reduced to $2^V$ for some $V \subset U$ by the ADD manipulation algorithms that
optimise the ADD representation of a function. For a simple example, consider
$\alpha: 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ defined as $\alpha(\{a\}) =
\alpha(\emptyset) = 0.5$. Then $\alpha$ can be reduced to $\alpha'\colon
2^{\emptyset} \to \mathbb{R}_{\ge 0}$ defined as $\alpha'(\emptyset) = 0.5$. To
compensate for these reductions, for the original WMC format with a weight
function $w\colon U \cup \{ \neg u \mid u \in U \} \to \mathbb{R}_{\ge 0}$,
ADDMC would multiply its computed answer by $\prod_{u \in U \setminus V} w(u) +
w(\neg u)$. With the new WMC format, we instead multiply the answer by $2^{|U
  \setminus V|}$. Each `excluded' variable $u \in U \setminus V$ satisfies two
properties: all weights associated with $u$ are equal to $0.5$ (otherwise the
corresponding CPT would depend on $u$, and $u$ would not be excluded), and all
other CPTs are independent of $u$ (or they may have a trivial dependence, where
the probability stays the same if $u$ is replaced with its complement). Thus,
the CPT that corresponds to $u$ still multiplies the weight of every atom in the
Boolean algebra by $0.5$, but the number of atoms being considered is halved. To
correct for this, we multiply the final answer by two for every $u \in U
\setminus V$.

@@

\paragraph{Observations.}
\begin{itemize}
\item The order of the encodings from best to worst is exactly the opposite of
  that in the previous literature.
\item Our encoding is particularly promising on instances of Grid networks.
  They are set up so that the entire Bayesian network is relevant to the
  query.
\item However, \texttt{db21} struggles with instances from Mastermind and
  Random Blocks domains (but so does \texttt{sbk05}). We conjecture that this
  is so either because the particular structure of these problems confuse the
  heuristics used by ADDMC or because these instances allow for significant
  simplifications that are exploited by $\mathtt{d02}$ but not
  $\mathtt{db21}$.
\item An observation from the cumulative plot: \texttt{db21} solves the same
  number of instances in \SI{6.374}{\second} as \texttt{d02} does in
  \SI{1000}{\second}.
\end{itemize}

@@

\paragraph{Explaining the performance benefits.} Let $n = |\mathcal{V}|$ be the
number of vertices in the Bayesian network, $d = \max_{X \in \mathcal{V}}
|\mathrm{pa}(X)|$ the maximum in-degree (i.e., number of parents), and $v =
\max_{X \in \mathcal{V}} |\im X|$ the maximum number of values per variable.
\Cref{tbl:asymptotes} shows how \texttt{db21} has both fewer variables and fewer
ADDs than any other encoding. However, note that these are upper bounds and most
encodings (including \texttt{db21}) can be smaller in certain situations (e.g.,
with binary random variables or when a CPT has repeating probabilities). We
equate clauses and ADDs (more specifically, factors of the function $\phi$ from
\cref{alg:encoding}) here because ADDMC interprets each clause of any WMC
encoding as a multiplicative factor of the ADD that represents the entire WMC
instance \cite{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each
weight is also a factor, but that has no effect on the asymptotic bounds in the
table.

>>>>

\begin{abstract}
  Weighted model counting (WMC) has emerged as the unifying inference mechanism
  across many (probabilistic) domains. Encoding an inference problem as an
  instance of WMC typically necessitates adding extra literals and clauses. This
  is partly so because the predominant definition of WMC assigns weights to
  models based on weights on literals, and this severely restricts what
  probability distributions can be represented. We propose a way to encode
  conditional weights on literals analogously to conditional probabilities. This
  representation can be as succinct as standard WMC with weights on literals but
  can also expand as needed to represent probability distributions with less
  structure. To demonstrate the performance benefits of conditional weights over
  the addition of extra literals, we develop a new WMC encoding for Bayesian
  networks and adapt a state-of-the-art WMC algorithm ADDMC to the new format.
  Our experiments show that the new encoding results in significantly faster
  inference on most benchmark instances. Furthermore, the same idea can
  be adapted to other WMC algorithms and other problem domains.
\end{abstract}

@@

\Cref{sec:prelims} describes three equivalent perspectives on the subject based
on logic, set theory, and Boolean algebras. Furthermore, we describe the space
of functions on Boolean algebras and various operations on those functions.
\Cref{sec:wmc_as_measure} introduces WMC as the problem of computing the value
of a measure on a Boolean algebra. We show that not all measures can be
represented using literal-based WMC, but all Boolean algebras can be extended to
make any measure representable in such a manner.

@@

Finally, in \cref{sec:experiments}, we show how the new encoding results in
significantly improved inference speed on most benchmark instances, often by one
or two orders of magnitude. We explain the performance benefits by showing how
the new encoding has asymptotically fewer variables and ADDs.

@@

\paragraph{ADDs and their use in probabilistic inference.} ADDs provide an
efficient way to manipulate functions from a Boolean algebra to any algebraic
structure (most commonly the real numbers)
\cite{DBLP:journals/fmsd/BaharFGHMPS97}. They have been used to represent value
functions in Markov decision processes \cite{DBLP:conf/uai/HoeySHB99} and, for
Bayesian network inference, to represent each conditional probability table
(CPT) as an ADD \cite{DBLP:conf/icml/ZhaoMP15} as well as by combining ADDs and
arithmetic circuits into a  single representation
\cite{DBLP:conf/ijcai/ChaviraD07}. ADDs are particularly advantageous in this
situation because of their ability to fully exploit context-specific
independence, i.e., the observable structure within a CPT that is not inherited
from the structure of the Bayesian network \cite{DBLP:conf/uai/BoutilierFGK96}.

@@

\subsection{The Space of Functions on Boolean Algebras}
% TODO: clarify why this is needed (and run a grammar check on this section)

@@

Note that in this case, we chose to interpret $\nu_b$ as $\Pr(b \mid a)$
while---with a different definition of $\nu_b$ that represents the joint
probability distribution $\Pr(a, b)$---$\nu_b$ by itself could induce $\mu$. In
general, however, factorising the full weight function into several smaller
functions often results in weight functions with smaller domains which leads to
increased efficiency and decreased memory usage \cite{DBLP:conf/aaai/DudekPV20}.

@@

Second, even though the function $\phi$ produced by \cref{alg:encoding} is
constructed to have $2^U$ as its domain, sometimes the domain is effectively
reduced to $2^V$ for some $V \subset U$ by the ADD manipulation algorithms that
optimise the ADD representation of a function. For a simple example, consider
$\alpha: 2^{\{a\}} \to \mathbb{R}_{\ge 0}$ defined as $\alpha(\{a\}) =
\alpha(\emptyset) = 0.5$. Then $\alpha$ can be reduced to $\alpha'\colon
2^{\emptyset} \to \mathbb{R}_{\ge 0}$ defined as $\alpha'(\emptyset) = 0.5$. To
compensate for these reductions, for the original WMC format with a weight
function $w\colon U \cup \{ \neg u \mid u \in U \} \to \mathbb{R}_{\ge 0}$,
ADDMC would multiply its computed answer by $\prod_{u \in U \setminus V} w(u) +
w(\neg u)$. With the new WMC format, we instead multiply the answer by $2^{|U
  \setminus V|}$. Each `excluded' variable $u \in U \setminus V$ satisfies two
properties: all weights associated with $u$ are equal to $0.5$ (otherwise the
corresponding CPT would depend on $u$, and $u$ would not be excluded), and all
other CPTs are independent of $u$ (or they may have a trivial dependence, where
the probability stays the same if $u$ is replaced with its complement). Thus,
the CPT that corresponds to $u$ still multiplies the weight of every atom in the
Boolean algebra by $0.5$, but the number of atoms under consideration is halved.
To correct for this, we multiply the final answer by two for every $u \in U
\setminus V$.

@@

% TODO: write. then run through a grammar checker.
\paragraph{Observations.}
\begin{itemize}
\item The order of the encodings from best to worst is exactly the opposite of
  that in the previous literature.
\item Our encoding is particularly promising on instances of Grid networks.
  They are set up so that the entire Bayesian network is relevant to the
  query.
\item However, \texttt{db21} struggles with instances from Mastermind and
  Random Blocks domains (but so does \texttt{sbk05}). We conjecture that this
  is so either because the particular structure of these problems confuse the
  heuristics used by ADDMC or because these instances allow for significant
  simplifications that are exploited by $\mathtt{d02}$ but not
  $\mathtt{db21}$.
\item An observation from the cumulative plot: \texttt{db21} solves the same
  number of instances in \SI{6.374}{\second} as \texttt{d02} does in
  \SI{1000}{\second}.
\end{itemize}

@@

\paragraph{Explaining the performance benefits.} Let $n = |\mathcal{V}|$ be the
number of vertices in the Bayesian network, $d = \max_{X \in \mathcal{V}}
|\mathrm{pa}(X)|$ the maximum in-degree (i.e., number of parents), and $v =
\max_{X \in \mathcal{V}} |\im X|$ the maximum number of values per variable.
\Cref{tbl:asymptotes} shows how \texttt{db21} has both fewer variables and fewer
ADDs than any other encoding. However, note that these are upper bounds and most
encodings (including \texttt{db21}) can be smaller in certain situations (e.g.,
with binary random variables or when a CPT has repeating probabilities). We
equate clauses and ADDs (more specifically, factors of the function $\phi$ from
\cref{alg:encoding}) here because ADDMC interprets each clause of any WMC
encoding as a multiplicative factor of the ADD that represents the entire WMC
instance \cite{DBLP:conf/aaai/DudekPV20}. For literal-weight encodings, each
weight is also a factor, but that does not affect the asymptotic bounds in the
table.
